---
title: "Day 3"
author: "Val Syverson, with updates by Jessica Blois"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# IMPORTANT NOTE
THIS TUTORIAL WAS USED FOR DEVELOPMENT ONLY, BUT NOT USED IN THE FINAL WORKSHOP. INSTEAD, STUDENTS DO A LIBRARY SEARCH AND PRESENTATION INSTEAD ON DAY 3

Today we'll explore further into what we can do with species distribution modeling.

# Part 1. Extending your model with land use data

Once again, load your workspace from yesterday: go to the Documents folder, choose the .Rdata file that you saved yesterday afternoon, and open it in RStudio. It should look exactly like it did yesterday (i.e. it should contain the objects from yesterday's models).

Run the package installation script again to load the packages you need.

```{r,eval = F}
source("installscript.R")
```

```{r,echo = F}
load(file = "Blois_Day2.RData")
```

```{r}
require(geodata);require(dismo);require(maps);require(here); require(sf); require(concaveman)
```


```{r}
sfdata <- st_as_sf(data_now,coords = c("lon","lat")) #this reformats the coordinate points into a special data structure
conc <- concaveman(sfdata,concavity = 3,length_threshold = 0) #this is called a concave hull, it's a polygon that contains all the points
conv <- convHull(data_now) #this is called a convex hull, it's just a polygon drawn around all the points that stick out the most
```

```{r}
xrange <- range(c(data_lgm$lon,data_midH$lon,data_now$lon),na.rm = T) + c(-5,5)
yrange <- range(c(data_lgm$lat,data_midH$lat,data_now$lat),na.rm = T) + c(-5,5)

plot(prednow, main = "today",xlim = xrange,ylim = yrange)
points(data_now$lon,data_now$lat,col = "maroon1",pch = 20,cex = 0.8)
plot(conv,add = T,col = rgb(0,0,0,0.3),lty = "blank")
plot(conc,add = T,col = rgb(0,1,1,0.3),lty = "blank")
```

```{r}
footprint <- footprint(year = "2009",path = tempdir())
```


# Part 2. Saving your data

Save your data. This is not straightforward on UC Merced computer lab computers, so please follow ALL of the following steps:

1.  Choose Session\>Save Workspace As...
2.  In the popup window, choose Documents from the list on the left side under Quick Access.
3.  Give the file a UNIQUE name with YOUR NAME in it and click Save. For example, a good file name would be: Blois_Day3.RData (but replace "Blois" with your name!)

# Part 3. Sharing your work

We will each do some additional research on our chosen species. Then, we will create a short presentation and share our model and our information with our classmates.

Information on the presentations can be found in the file: 'student presentation template.pptx'

To complete the presentation, you may need to reload your saved workspace from Day 2 and Day 3 and 


