---
title: "Day 2"
author: "Val Syverson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1. Mathematical modeling

First make sure your variables from yesterday are loaded.
```{r}
load(file="day1.RData")
```

We're going to use the `bioclim` climatic data and the `data` file you constructed to model the influence of the climate variables on your species' range. 

First, get the value of each of the 19 climate variables at the location of each of the species occurrences with the function `extract` from the `raster` package. This function takes two arguments, a RasterStack and a set of coordinates (GPS points), and returns 
```{r}
presvals<-extract(bioclim, data[,c("lon","lat")])
dim(presvals)
head(presvals)
```

You can see that the dimension
```{r}
set.seed(0)
backgr <- randomPoints(bioclim, min(c(nrow(data),500)))
absvals <- extract(bioclim, backgr)
pb <- c(rep(1, nrow(presvals)), rep(0, nrow(absvals)))
sdmdata <- data.frame(cbind(pb, rbind(presvals, absvals)))
```


